Unsupervised embeddings of organizations capture the latent structure of global scientific mobility

### 2020-07-03 paper/current/main.tex
remove math section

The functional distance derived from the embedding explains and predicts scientific mobility better than geographic distance.
The amount of variance explained by embedding distance ($R^{2} = 0.46$, Fig.~\ref{fig:gravity_performance}a) is more than twice that of geographic distance ($R^{2} = 0.22$, Fig.~\ref{fig:gravity_performance}b).
This pattern holds when considering the subset of only domestic (Fig.~\ref{fig:supp:gravity_correlation}c) and only international mobility flows (Fig.~\ref{fig:supp:gravity_correlation}d).
After fitting model parameters from the data, the embedding distance produces better predictions of expected flux than does geographic distance (Fig.~\ref{fig:gravity_performance}c--d), which also holds for the subsets of only domestic (Fig.~\ref{fig:supp:gravity_correlation}e--f) and international mobility(Fig.~\ref{fig:supp:gravity_correlation}g--h).
The embedding distance defined from the cosine distnace between vectors also outperforms that defined from their dot product ($R^{2} = 0.34$, Fig.~\ref{fig:supp:gravity_dot}).
The embedding distance also out-performs alternative ``diffusion'' measures such as the personalized-page rank scores (see supporting materials) calculated from the underlying mobility network (Fig.~\ref{fig:supp:gravity_pprcos}, Fig.~\ref{fig:supp:gravity_pprjsd}).
Geographic distance alone is a poor perspective for understanding international scientific mobility---rather, by learning a functional distance directly from the data, it is possible to better explain, and better predict flux between organizations.

@@

%
% Subsection: Mathematical Exploration
%
\subsection*{Mathematical connection between gravity model and neural embedding }

Why does the neural embedding model (skip-gram with negative sampling, or \textit{word2vec}), originally designed for natural language processing~\autocite{mikolov2013word2vec}, produce a reasonable distance that models gravity relationships between organizations?
We find evidence of a theoretical connection between the two models, one that makes the neural embedding model applicable to embedding many forms of mobility trajectories. The gravity model generates the flow from $i$ to $j$ by

\begin{align}
    \label{eq:gravity_model}
    T_{ij}:= G M_i M_j f\left(d_{ij}\right),
\end{align}

where $M_i$ and $M_j$ are the masses of of the entities, $f\left(d_{ij}\right)$ is a function of the distance, and $G$ is the constant analogous to the gravitational constant in physics. 
In mobility studies, any quantity proportional to the sum of flows from $i$ can be used as $M_i$; a common choice is the population at a point $i$ (e.g., city and country).
For $d_{ij}$, geographical distance is often used. 
We argue that the neural embedding model
Here, we show that the neural embedding model can be interpreted as generalized version of the gravity model that efficiently encodes a functional distance into the dense, continuous, and meaningful embedding vector.

In the case of language, the neural embedding model attempts to maximize the average log probability of observing a target word in a sentence, $w_{t}$, given the words in the surrounding context window, $w_{t + k}$. 
This can be written as follows, where $c$ is the size of the context window and $T$ iterates through possible target words:

\begin{equation}
    \label{eq:skipgram_log_prob_condensed}
    \frac{1}{T}\sum_{t=1}^{T} \sum_{-c \le j \le c, j \neq 0} \log p(w_{t+j}|w_t),
\end{equation}

where the conditional probability of observing one word $i$, given word, $j$, is defined as,

\begin{align}
  \label{eq:model_node2vec}
    P \left(j \mid i\right) = \frac{1}{Z_i}\exp\left(\vec{u} ^\top _{i} \vec{v}_{j}\right), 
\end{align}

where $u_i$ and $v_j$ are `the `in-vector'' and ``out-vector'' respectively, which are K-dimensional vector representations of each word in the vocabulary. 
Variable $Z_i$ is the normalization constant or the partition function, i.e., $Z_i:= \sum_{\ell=1}^N \exp\left(\vec{u} ^\top _{i} \vec{v}_{\ell}\right)$.  By the definition of the euclidean distance between the in-vectors,

\begin{align}
    \label{eq:distance}
    d(\vec{u},\vec{u}'):= \sqrt{\sum_{k=1}^K \left(u_{k} - u'_{k}\right)^2}.
\end{align}

For simplicity, we assume that  $\vec{v}_j = \vec{u}_j$, although they may differ in practice.
Exploiting the fact that $d^2 (\vec{u},\vec{u}')=||\vec{u}||^2 + ||\vec{u}'||^2 - 2\vec{u} ^\top \vec{u}'$, the dot similarly between in and our vectors $\vec{u}_i$ and $\vec{v}_j$ is given by

\begin{align}
    \label{eq:dotsim}
    \vec{u}_i ^\top \vec{v}_j = \frac{1}{2}\left[ ||\vec{u}_i||^2 + ||\vec{u}_j||^2 - d\left(\vec{u}_i, \vec{u}_j\right)^2 \right],
\end{align}

Using Eq.~\ref{eq:dotsim}, we can define $\exp\left( \vec{u}_i ^\top \vec{v}_j \right)$ as

\begin{align}
    \exp\left( \vec{u}_i ^\top \vec{v}_j \right) &= \frac{\exp\left( ||\vec{u}_i||^2/2 + ||\vec{u}_j||^2/2 \right)}{\exp\left[ d\left(\vec{u}_i, \vec{u}_j\right)^2 / 2 \right]} \nonumber \\
                                                 &= \frac{s_i s_j}{\exp\left[d\left(\vec{u}_i, \vec{u}_j\right)^2 / 2\right]}, \label{eq:exp_distance_form}
\end{align}

where $s_i:=\exp(||\vec{u} _i||^2/2)$. 
Substituting Eq.~\eqref{eq:exp_distance_form} into Eq.~\eqref{eq:model_node2vec} yields

\begin{align}
    \label{eq:model_node2vec_distance}
    P \left(j \mid i\right) &:= \frac{1}{Z_i}\cdot \frac{s_is_j}{\exp\left[d\left(\vec{u}_i, \vec{u}_j\right)^2/2\right]}. 
\end{align}

If we consider a stable Markov process with the transition probability from $i$ to $j$, $P(j \vert i)$ and consider the stationary distribution for the Markov process  $\pi$, i.e., mobile scientists. Then, the flow from $i$ to $j$  in one step is given by

\begin{align}
    \label{eq:flow}
    T_{ij}:= \pi_i P(j \vert i).
\end{align}

According to detailed balanced condition of Markov process,

\begin{align}
    \pi_i P\left(j \vert i \right) = \pi_j P\left(i \vert j \right),
\end{align}

\begin{align}
    \label{eq:detailed_balanced}
    &\frac{\pi_i}{Z_i}\cdot \frac{s_i s_j}{\exp\left[d\left(\vec{u}_i, \vec{u}_j\right)^2 / 2\right]} = \frac{\pi_j}{Z_j}\cdot \frac{s_i s_j}{\exp\left[d\left(\vec{u}_i, \vec{u}_j\right)^2 / 2 \right]} \\
    &\iff \frac{\pi_i}{Z_i} = \frac{\pi_j}{Z_j} =: G', \label{eq:const}
\end{align}

where $G'$ is a constant. 
Substituting Eq.~\eqref{eq:model_node2vec_distance} into Eq.~\eqref{eq:flow} and exploiting Eq.~\eqref{eq:const} yield

\begin{align}
    \label{eq:flow_gravity_model}
    T_{ij} &= \pi_i P\left(j \vert i \right)\nonumber \\
           &= \frac{\pi_i}{Z_i} \cdot \frac{s_i s_j}{\exp\left[d\left(\vec{u}_i, \vec{u}_j\right)^2 / 2 \right]}\nonumber \\ 
           &= G' \frac{s_i s_j}{\exp\left[d\left(\vec{u}_i, \vec{u}_j\right)^2 / 2 \right]}. 
\end{align}

The form of Eq.~\ref{eq:flow_gravity_model} maps to the gravity model $T_{ij}:= G M_i M_j f\left(d_{ij}\right)$.
Therefore, we argue that the nerual embedding model is a generalized version of the gravity model with $M_i=s_i$ and $f\left(d_{ij} \right) = \exp\left[ - d\left(\vec{u}_i,\vec{u}_j\right) ^ 2 /2\right]$.
Given flows $T_{ij}$ derived from mobility trajectories, this model attempts to learn the mass $M_i$ and map each entity into more abundant vector space . 
Hence, the neural embedding model can be an effective model for investigating any kind of mobility trajectory data, stemming from the mathematical relationship between it and the gravity model of mobility.

>>>>

The functional distance derived from the embedding explains and predicts scientific mobility better than geographic distance.
The amount of variance explained by embedding distance ($R^{2} = 0.46$, Fig.~\ref{fig:gravity_performance}a) is more than twice that of geographic distance ($R^{2} = 0.22$, Fig.~\ref{fig:gravity_performance}b).
This pattern holds when considering the subset of only domestic (Fig.~\ref{fig:supp:gravity_correlation}c) and only international mobility flows (Fig.~\ref{fig:supp:gravity_correlation}d).
After fitting model parameters from the data, the embedding distance produces better predictions of expected flux than does geographic distance (Fig.~\ref{fig:gravity_performance}c--d), which also holds for the subsets of only domestic (Fig.~\ref{fig:supp:gravity_correlation}e--f) and international mobility(Fig.~\ref{fig:supp:gravity_correlation}g--h).
The embedding distance defined from the cosine distnace between vectors also outperforms that defined from their dot product ($R^{2} = 0.34$, Fig.~\ref{fig:supp:gravity_dot}).
The embedding distance also out-performs alternative ``diffusion'' measures such as the personalized-page rank scores (see supporting materials) calculated from the underlying mobility network (Fig.~\ref{fig:supp:gravity_pprcos}, Fig.~\ref{fig:supp:gravity_pprjsd}).
Geographic distance alone is a poor perspective for understanding international scientific mobility---rather, by learning a functional distance directly from the data, it is possible to better explain, and better predict the flow of scientific mobility between organizations.
We speculate that this results from an underlying mathematical connection between the neural embedding architecture and the gravity model of mobility.

@@
